\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-04-24 doc/incertae_sedis.tex
Tiny edits.

Supertree methods combine a collection of input trees with different
taxon sets into a single tree on the combined taxon set.
The methods are usually applied to leaf-labeled phylogenetic estimates. 
Here we describe a supertree approach that overcomes issues associated with
using a taxonomy as one of the inputs to a supertree method.

@@

The supertree method described by \citet{redelings2017supertree} is

able to use a set of phylogenetic estimates and taxonomy to produce a
supertree with clades names according to the taxonomy.
However, the
method was not able to accommodate the fact that taxonomies frequently
contain nodes with uncertain placement.
These nodes are often labelled
``\emph{incertae sedis}'', which means ``uncertain seat'' in Latin.
The common interpretation of such taxa is that they may not be moved
further towards the root, but may be moved into their sibling taxa.
For example, a genus with a sibling that is a family may be annotated
as ``\emph{incertae familia}'', indicating that it is unclear which
family the genus should be placed in.
\emph{Incertae sedis} taxa
frequently occur when a specimen is identified down to a given
taxonomic level, but no further.
Extinct taxa are often \emph{incertae
sedis}.
OTT is constructed from several source taxonomies, and these
sources include various ways of indicating that the position of a
taxon within the hierarchy is uncertain.
OTT uses a series of
``flags'' to annotate these taxa.
The supertree of by
\citet{redelings2017supertree} simply pruned these taxa from the
taxonomy and input trees.
Thus, the final supertree lacked taxa which
were \emph{incertae sedis} or the descendants of such taxa.

Here describe a supertree method so that we can resolve taxonomic
uncertainty by using published phylogenies to place \emph{incertae
sedis} taxa.
Developing this method required deciding on a set
operational semantics to be used when interpreting the \emph{incertae
sedis} annotation.
These semantics affect the interpretation of the
phylogenetic statements being made by the input taxonomy and the rules
for applying taxonomic names to the final supertree.
In addition to
discussing how to interpret the \emph{incertae sedis} annotation,
we describe the algorithmic changes to the supertree pipeline that
were required to adequately represent the taxonomic uncertainty.
The
approach described here allows us to include thousands of new taxa in
our supertree analysis that were previously filtered out.
It will also
enable the Open Tree of Life project to include extinct taxa, since
many of these taxa are \emph{incertae sedis}.
This makes substantial
progress towards our goal of comprehensive inclusion of known taxa
into the Open Tree of Life summary tree.

@@

The supertree algorithm of \citet{redelings2017supertree} takes as
input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of leaf-labeled rooted input phylogenies and a single rooted taxonomy
tree $\taxonomy_{p}$ that is ranked below all of the input trees.
As
discussed in \citet{redelings2017supertree}, the preferred tree would
display as many of the highly ranked input splits as possible.
If
computational barriers were not an issue, the method would act by
first producing a ranked list of rooted splits:
$\mathcal{S}=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$ where
$S(T_{i})$ denotes a list of non-trivial rooted splits created from a
post-order traversal of tree $i$ and '$+$' denotes concatenation.
Then
the supertree would be produced in a greedy manner by iterating over
the list $S$, and adding each rooted split to a growing rooted
supertree if that split is compatible with the supertree.
In other
words, the set $C$ is initialized to the empty set, and we then
iteratively consider each split in the list $\mathcal{S}$ and insert
it into $C$ if the resulting set remains jointly compatible.
The BUILD
algorithm\citep{AhoSSU1981} is used both to check compatibility, and
to construct a compatible tree from the final set $C$.

@@

The taxonomy tree, $\taxonomy_{p}$, is derived from OTT.
As mentioned
above, OTT not only contains a taxonomic hierarchy, but also contains
taxonomic labels and annotations (referred to as ``flags'') on the
taxa to carry extra information.
The entire hiearchy encoded by OTT is
a complete taxonomic tree, $\taxonomy^{\ast}.$ Nodes in
$\taxonomy^{\ast}$ have a one-to-one correspondence with a set of
labels $\mathcal{N}$ that are called taxon names.
We will decompose
$\mathcal{N}$ into the taxon names for leaf taxa \emph{$\mathcal{L}$
}and higher taxa $\mathcal{H}$.
We will use $\mathcal{U}$ to refer to
the subset ($\mathcal{U}\subseteq\mathcal{N}$) of taxa that are
annotated with the \emph{incertae sedis }property.
Other annotations
identify dubious taxa and taxonomic names that are unwanted artifacts
of the process of building OTT.
Thus, the taxonomic tree which is used
as an input to the supertree algorithm is pruned version of
$\taxonomy^{\ast}$.
For our purposes, the relevant information from
the taxonomy consists of the taxonomy tree $\taxonomy^{\ast}$, the
labels $\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis} taxa.
In previous work, taxa in $\mathcal{U}$ and all of their descendants 
were all pruned from
the taxonomic hierarchy when $\taxonomy_{p}$ is extracted from OTT.
In
the present work, we consider supertree approaches that operate on a
taxonomic tree $\taxonomy$, which is produced from
$\taxonomy^{\ast}$but does not prune \emph{incertae sedis} taxa.

Throughout we assume that leaf labels on input phylogenetic trees are
only taken from the leaf set of the taxonomic tree that is being used
and that each label occurs at most once in each input tree.
\citet{redelings2017supertree} describe a method called
``exemplification'' that ensures this is the case.
Thus pruning of the
\emph{incertae sedis} taxa from the taxonomy was also accompanied by
pruning those taxa from the input phylogenetic trees.

@@

In terms of phylogenetic nomenclature the supertree pipeline acts as
    the taxonomy defines a branch-based name \citep[see][]{deQueiroz2013}
    for each higher taxon of the form.
So for name labeling node $n$ in the taxonomic tree, 
    the operational definition would be:
    ``the clade defined by all taxa
    that are more closely related to $\mathcal{I}(n)$ than to
    $\mathcal{E}(n)$, if such a clade exists on the tree.''

>>>>

Supertree methods combine a collection of input trees with different
taxon sets into a single tree on the combined taxon set.
These methods are usually applied to leaf-labeled phylogenetic estimates. 
Here we describe a supertree approach that overcomes issues associated with
using a taxonomy as one of the inputs to a supertree method.

@@

The supertree method described by \citet{redelings2017supertree} is
able to use a set of phylogenetic estimates and taxonomy to produce a
supertree with clades names according to the taxonomy.
However, the
method was not able to accommodate the fact that taxonomies frequently
contain nodes with uncertain placement.
These nodes are often labelled
``\emph{incertae sedis}'', which means ``uncertain seat'' in Latin.
The common interpretation of such taxa is that they may not be moved
further towards the root, but may be moved into their sibling taxa.
For example, a genus with a sibling that is a family may be annotated
as ``\emph{incertae familia}'', indicating that it is unclear which
family the genus should be placed in.
\emph{Incertae sedis} taxa
frequently occur when a specimen is identified down to a given
taxonomic level, but no further.
Extinct taxa are often \emph{incertae
sedis}.
OTT is constructed from several source taxonomies, and these
sources include various ways of indicating that the position of a
taxon within the hierarchy is uncertain.
OTT uses a series of
``flags'' to annotate these taxa.
The supertree of \citet{redelings2017supertree} simply pruned these taxa from the
taxonomy and input trees.
Thus, the final supertree lacked taxa which
were \emph{incertae sedis} or the descendants of such taxa.

Here describe a supertree method so that we can resolve taxonomic
uncertainty by using published phylogenies to place \emph{incertae
sedis} taxa.
Developing this method required deciding on a set of
operational semantics to be used when interpreting the \emph{incertae
sedis} annotation.
These semantics affect the interpretation of the
phylogenetic statements being made by the input taxonomy and the rules
for applying taxonomic names to the final supertree.
In addition to
discussing how to interpret the \emph{incertae sedis} annotation,
we describe the algorithmic changes to the supertree pipeline that
were required to adequately represent the taxonomic uncertainty.
The
approach described here allows us to include thousands of new taxa in
our supertree analysis that were previously filtered out.
It will also
enable the Open Tree of Life project to include extinct taxa, since
many of these taxa are \emph{incertae sedis}.
This makes substantial
progress towards our goal of comprehensive inclusion of known taxa
into the Open Tree of Life summary tree.

@@

The supertree algorithm of \citet{redelings2017supertree} takes as
input a ranked list $\phyloinputs=\left\{ T_{1},\ldots,T_{n}\right\} $
of leaf-labeled rooted input phylogenies and a single rooted taxonomy
tree $\taxonomy_{p}$ that is ranked below all of the input trees.
As
discussed in \citet{redelings2017supertree}, the preferred tree would
display as many of the highly ranked input splits as possible.
If
computational barriers were not an issue, the method would act by
first producing a ranked list of rooted splits:
$\mathcal{S}=S(T_{1})+S(T_{2})+\ldots+S(T_{n})+S(\mathbb{T})$ where
$S(T_{i})$ denotes a list of non-trivial rooted splits created from a
post-order traversal of tree $i$ and '$+$' denotes concatenation.
Then
the supertree would be produced in a greedy manner by iterating over
the list $S$, and adding each rooted split to a growing rooted
supertree if that split is compatible with the supertree.
In other
words, the set $C$ is initialized to the empty set, and we then
iteratively consider each split in the list $\mathcal{S}$ and insert
it into $C$ if the resulting set remains jointly compatible.
The BUILD
algorithm \citep{AhoSSU1981} is used both to check compatibility, and
to construct a compatible tree from the final set $C$.

@@

The taxonomy tree, $\taxonomy_{p}$, is derived from OTT.
As mentioned
above, OTT not only contains a taxonomic hierarchy, but also contains
taxonomic labels and annotations (referred to as ``flags'') on the
taxa to carry extra information.
The entire hierarchy encoded by OTT is
a complete taxonomic tree, $\taxonomy^{\ast}.$ Nodes in
$\taxonomy^{\ast}$ have a one-to-one correspondence with a set of
labels $\mathcal{N}$ that are called taxon names.
We will decompose
$\mathcal{N}$ into the taxon names for leaf taxa \emph{$\mathcal{L}$
}and higher taxa $\mathcal{H}$.
We will use $\mathcal{U}$ to refer to
the subset ($\mathcal{U}\subseteq\mathcal{N}$) of taxa that are
annotated with the \emph{incertae sedis }property.
Other annotations
identify dubious taxa and taxonomic names that are unwanted artifacts
of the process of building OTT.
Thus, the taxonomic tree which is used
as an input to the supertree algorithm is pruned version of
$\taxonomy^{\ast}$.
For our purposes, the relevant information from
the taxonomy consists of the taxonomy tree $\taxonomy^{\ast}$, the
labels $\mathcal{N}$, and set $\mathcal{U}$ of \emph{incertae sedis} taxa.
In previous work, taxa in $\mathcal{U}$ and all of their descendants 
were all pruned from
the taxonomic hierarchy when $\taxonomy_{p}$ is extracted from OTT.
In
the present work, we consider supertree approaches that operate on a
taxonomic tree $\taxonomy$, which is produced from
$\taxonomy^{\ast}$but does not prune \emph{incertae sedis} taxa.

Throughout we assume that leaf labels on input phylogenetic trees are
only taken from the leaf set of the taxonomic tree that is being used
and that each label occurs at most once in each input tree.
\citet{redelings2017supertree} describe a method called
``exemplification'' that ensures this is the case.
Thus pruning of the
\emph{incertae sedis} taxa from the taxonomy was also accompanied by
pruning those taxa and all of their descendants from the input phylogenetic
trees.

@@

In terms of phylogenetic nomenclature the supertree pipeline acts as
if the taxonomy defines a branch-based name \citep[see][]{deQueiroz2013}
    for each higher taxon of the form.
So for the name labeling node $n$ in the taxonomic tree, 
    the operational definition would be:
    ``the clade defined by all taxa
    that are more closely related to $\mathcal{I}(n)$ than to
    $\mathcal{E}(n)$, if such a clade exists on the tree.''
